---
title: "Storm Events with Respect to Health & Economic Consequences ( U.S.A. - 1950 to 2011)"
author: "P.Y.Fong"
date: "17 Dec 2020"
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
library (knitr)
opts_chunk$set(cache =T,
               echo = T,
               message=F,
               warning=F)
```

# Synopsis
This report seeks to answer 2 questions:  

1.    Across the United States, which types of weather events are most harmful with respect to population health?
2.    Across the United States, which types of weather events have the greatest economic consequences?

# Library & Data Loading
The data used in this analysis was based on data from the link provided : [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) (47Mb)  

The data is a copy of the StormData.CSV file recorded between 1950 and Nov 2011 originally sourced from the US National Weather Service.  

We first create a download sub directory and download the file from the URL into it on the first run only.  
The various libraries required for analysis using R is also loaded.

```{r libraries, download}
library(data.table)
library(R.utils)
library(dplyr)
library(ggplot2)
library(grid)
library(gridExtra)

# Download source files from URL into and create download sub-directory on 1st run only.
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
destfile <- "./download/StormData.csv.bz2"

if(!file.exists("./download/StormData.csv.bz2")) {
  dir.create("./download")  
  download.file(url,destfile=destfile)
}
```

The data was then read from the file into a data table.  

```{r data load}
data <- fread("./download/StormData.csv.bz2")
```
Examining the data we find the following:
```{r exam data}
str(data)
```

## Selecting & Cleaning Data
The data set to be examined to answer the above questions only required the following variables:  
* BGN_DATE    (CHR)  
* EVTYPE      (CHR)  
* FATALITIES  (NUM)  
* INJURIES    (NUM)  
* PROPDMG     (NUM)  
* PROPDMGEXP  (CHR)  
* CROPDMG     (NUM)  
* CROPDMGEXP  (CHR)  

```{r Select variables}
# select required variables
data <- data %>% select(EVTYPE,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)
```

Checking for missing values.  

```{r missing values}
mean(is.na(data))
```

Further examination of the selected data was performed.  
The following was noted:  
1. The variables order of magnitude of property damage, PROPDMGEXP and crop damage, CROPDMGEXP had more unique values than stated in the documentation.  
* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)  
* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)  
i.e. (H= hundreds, K= thousands, M= millions, B= billions).

```{r Check Key Variables}
# Checking prop dmg. exp
# Checking prop dmg. exp
rd <- 4
PDE <- data %>%
  count(PROPDMGEXP, name='PDEcount') %>%
  mutate (PDE_pct = round(100*PDEcount/sum(PDEcount),rd))
#t1 <- tableGrob(PDE, rows = NULL, cols = c("Prop.Dmg.Exp.","Count","%"))

# Checking crop dmg. exp
CDE <- data %>%
  count(CROPDMGEXP, name='CDEcount') %>%
  mutate (CDE_pct = round(100*CDEcount/sum(CDEcount),rd))
#t2 <- tableGrob(CDE, rows = NULL, cols = c("Crop.Dmg.Exp.","Count","%"))

# Printing tables of unique values
kable(list(PDE,CDE), format='pipe', format.args= list(big.mark = ","),
      caption = "Unique values in Property & Crop Damage EXP")
```

Although the erroneous EXP values for both EXP components were hardly significant it still had to be cleaned up.  

The erroneous EXP values were converted according to the following assumptions in order:  

1. All values were capitalized.  
2. All special characters were converted to *blanks*.  
3. All *blanks* were then converted to 0.  
4. Digits were considered as 1e+^(digit)^ multipliers.  

Property & crop damage values were then calculated and combined into variables "PropDmg" & "CropDmg" respectively in the  clean data set.

```{r Clean Data}
# Replace erroneous values of Property & Crop Damages EXP.

## Cleaning prop & crop dmg. exp
data$PROPDMGEXP <- gsub("[^[:alnum:]]","",data$PROPDMGEXP)
data$CROPDMGEXP <- gsub("[^[:alnum:]]","",data$CROPDMGEXP)
data <-data %>%
  mutate(PROPDMGEXP = toupper(PROPDMGEXP),CROPDMGEXP = toupper(CROPDMGEXP)) %>%
  mutate(PROPDMGEXP = replace(PROPDMGEXP,PROPDMGEXP=="",0),
         CROPDMGEXP = replace(CROPDMGEXP,CROPDMGEXP=="",0)) %>%
  mutate(PROPDMGEXP = replace(PROPDMGEXP,PROPDMGEXP=='H', 100),
         CROPDMGEXP = replace(CROPDMGEXP,CROPDMGEXP=='H', 100)) %>%
  mutate(PROPDMGEXP = replace(PROPDMGEXP,PROPDMGEXP=='K', 1000),
         CROPDMGEXP = replace(CROPDMGEXP,CROPDMGEXP=='K', 1000)) %>%
  mutate(PROPDMGEXP = replace(PROPDMGEXP,PROPDMGEXP=='M', 1000000),
         CROPDMGEXP = replace(CROPDMGEXP,CROPDMGEXP=='M', 1000000)) %>%
  mutate(PROPDMGEXP = replace(PROPDMGEXP,PROPDMGEXP=='B', 1000000000),
         CROPDMGEXP = replace(CROPDMGEXP,CROPDMGEXP=='B', 1000000000)) %>%
  mutate(PROPDMGEXP = as.double(PROPDMGEXP),CROPDMGEXP = as.double(CROPDMGEXP))

## Calculating Damage Values and combining into new variables.
cleandata <-  data %>%
  mutate(PropDmg = PROPDMG*PROPDMGEXP, CropDmg = CROPDMG*CROPDMGEXP) %>%
  select("EVTYPE", "FATALITIES", "INJURIES", "PropDmg", "CropDmg" )
head(cleandata)
```
## Analysing the data and results

The clean data set was filtered and grouped to show the required summaries.
Only the top 5 events in each category were plotted accordingly.  

Initial options for the plots:

```{r Plot Options}
# Options for Figures
mytheme <- ttheme_default(base_size=10,
                          core= list(fg_params= list(hjust=0,x=0.1)),
                          colhead= list(fg_params=list(hjust=0, x=0.1)))
top <-5
```

### Top 5 Weather Events Causing Fatalities.

```{r Fatalities}
## Fatalities
fatal <- cleandata %>%
  group_by(EVTYPE) %>%
  summarise(totfatal = sum(FATALITIES)) %>%
  arrange(desc(totfatal)) %>%
  filter(totfatal !=0) %>%
  mutate(pct=round(100*totfatal/sum(totfatal),1)) %>%
  slice_max(totfatal, n=top)

p1f <- ggplot(fatal,aes(x=totfatal, y=reorder(EVTYPE, totfatal)))
p1f <- p1f + geom_col(fill ='red')
p1f <- p1f + labs(title= paste('Fatalities by Event Type (Top',top,
                               '=', sum(fatal$pct),'%)'), y=NULL, x= 'Count')
fatal$totfatal <- format(fatal$totfatal, big.mark=",",core.just= "right")
tb1 <- tableGrob(fatal, theme= mytheme, rows = NULL, cols= c("Event Type","Fatalities","%"))

grid.arrange(p1f, tb1, nrow= 2, as.table = T, heights = c(3,2))
```

### Top 5 Weather Events Causing Injuries.

```{r Injured}
## Injured
injured <- cleandata %>%
  group_by(EVTYPE) %>%
  summarise(totinjured = sum(INJURIES)) %>%
  arrange(desc(totinjured)) %>%
  filter(totinjured !=0) %>%
  mutate(pct=round(100*totinjured/sum(totinjured),1)) %>%
  slice_max(totinjured,n=top)

p2i <- ggplot(injured, aes(y=reorder(EVTYPE, totinjured), x=totinjured))
p2i <- p2i + geom_col(fill ='orange')
p2i <- p2i + labs(title= paste('Injured by Event Type (Top',top,
                               '=', sum(injured$pct),'%)'), y=NULL, x= 'Count')
injured$totinjured <- format(injured$totinjured, big.mark=",",core.just= "right")
tb2 <- tableGrob(injured, theme= mytheme, rows = NULL, cols= c("Event Type","Injured","%"))
grid.arrange(p2i, tb2, nrow= 2, as.table = T, heights = c(3,2))
```

### Top 5 Weather Events Causing Property Damage.  

```{r PropDmg}
PDmg <- cleandata %>%
  group_by(EVTYPE) %>%
  summarise(totPDmg = round(sum(PropDmg)/1e9,3)) %>%
  arrange(desc(totPDmg)) %>%
  filter(totPDmg !=0) %>%
  mutate(pct=round(100*totPDmg/sum(totPDmg),1)) %>%
  slice_max(totPDmg,n=top)

p3p <- ggplot(PDmg, aes(y=reorder(EVTYPE, totPDmg), x= totPDmg))
p3p <- p3p + geom_col(fill ='Purple')
p3p <- p3p + labs(title= paste('Property Damage Value by Event Type (Top',top,
                               '=', sum(PDmg$pct),'%)'), y= NULL, x= 'Value ($B)')
PDmg$totPDmg <- format(PDmg$totPDmg, big.mark=",",core.just= "right")
tb3 <- tableGrob(PDmg, theme= mytheme, rows = NULL, cols= c("Event Type","Value ($B)","%"))
grid.arrange(p3p, tb3, nrow= 2, as.table = T, heights = c(3,2))
```

### Top 5 Weather Events Causing Crop Damage.  

```{r Crop Damage}
## Crop Damage
CDmg <- cleandata %>%
  group_by(EVTYPE) %>%
  summarise(totCDmg = round(sum(CropDmg)/1e9,3)) %>%
  arrange(desc(totCDmg)) %>%
  filter(totCDmg !=0) %>%
  mutate(pct=round(100*totCDmg/sum(totCDmg),1)) %>%
  slice_max(totCDmg,n=top)

p4c <- ggplot(CDmg,aes(y=reorder(EVTYPE, totCDmg), x=totCDmg))
p4c <- p4c + geom_col(fill ='Steel Blue')
p4c <- p4c + labs(title= paste('Crop Damage Value by Event Type (Top',top,
                               '=', sum(CDmg$pct),'%)'), y= NULL, x= 'Value ($B)')
CDmg$totCDmg <- format(CDmg$totCDmg, big.mark=",",core.just= "right")
tb4 <- tableGrob(CDmg, theme= mytheme, rows = NULL, cols= c("Event Type","Value ($B)","%"))
grid.arrange(p4c, tb4, nrow= 2, as.table = T, heights = c(3,2))
```
