---
title: "Weather Events with Respect to Health & Economic Consequences ( U.S.A. - 1950 to 2011)"
author: "P.Y.Fong"
date: "17 Dec 2020"
output:
  html_document:
    keep_md: true
    fig_width: 10 
    fig_height: 10 
---

```{r setup, include=FALSE}
library (knitr)
opts_chunk$set(cache =T,
               echo = T,
               message=F,
               warning=F)
```

# Synopsis
This report seeks to answer 2 questions:  

1.    Across the United States, which types of weather events are most harmful with respect to population health?
2.    Across the United States, which types of weather events have the greatest economic consequences?

# R Libraries
various libraries required for analysis using R were loaded.

```{r libraries}
library(data.table)
library(R.utils)
library(dplyr)
library(ggplot2)
library(grid)
library(gridExtra)
```

# Data Source & Download
The data used in this analysis was based on data from the link provided : [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) (47Mb)  

The data is a copy of the StormData.CSV file recorded between 1950 and Nov 2011 originally sourced from the US National Weather Service.  

We first create a download sub directory and download the file from the URL into it on the first run only.  

Documentation on the data is available from these links:  
* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)  
* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)  

```{r download}
# Download source files from URL into and create download sub-directory on 1st run only.
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
destfile <- "./download/StormData.csv.bz2"

if(!file.exists("./download/StormData.csv.bz2")) {
  dir.create("./download")  
  download.file(url,destfile=destfile)
}
```

## Data Processing

The data was then read from the file into a data table.  

```{r data load}
data <- fread("./download/StormData.csv.bz2")
```
Examining the data we find the following:
```{r exam data}
str(data)
```

The data set to be examined to answer the above questions only required selecting the following variables:  
* EVTYPE      (CHR)  
* FATALITIES  (NUM)  
* INJURIES    (NUM)  
* PROPDMG     (NUM)  
* PROPDMGEXP  (CHR)  
* CROPDMG     (NUM)  
* CROPDMGEXP  (CHR)  

```{r Select variables}
# select required variables
data <- data %>% select(EVTYPE,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)
```

The data was found to have no clear NA missing values.  

```{r missing values}
mean(is.na(data))
```

Further examination of the selected data was performed.  

### Examining Variable EVTYPE

This variable had an extremely large variance in values as compared to the 48 types documented.  

```{r Check EVTYPE}
# Check unique variables in EVTYPE
length(unique(data$EVTYPE))
```

Only minor cleaning was performed on this variable which were in order:  

1. All values were capitalized.  
2. Leading, Trailing and double spaces were removed.  

```{r Clean EVTYPE}
# Cleaning EVTYPE
data$EVTYPE <- toupper(data$EVTYPE)
data$EVTYPE <- trimws(data$EVTYPE)
searchString <- '  '
replacementString <- ' '
data$EVTYPE = sub(searchString,replacementString,data$EVTYPE)

# Check unique variables in EVTYPE
length(unique(data$EVTYPE))
```
This somewhat reduced the unique values.   

Notes:  
1. The Weather event types in the data was highly contextual and any matching / correction to the documented types required weather services expertise.  
2. It was targeted that the top 5 event types should represent at least 60% of the total from the data set or else further cleaning would be required.  

### Examining Damage Order of Magnitude Variables

The order of magnitude variables for property damage, PROPDMGEXP and crop damage, CROPDMGEXP had more unique values than stated in the documentation. i.e. (H= hundreds, K= thousands, M= millions, B= billions).  

```{r Check EXP Variables}
# Checking prop dmg. exp
rd <- 4
PDE <- data %>%
  count(PROPDMGEXP, name='PDEcount') %>%
  mutate (PDE_pct = round(100*PDEcount/sum(PDEcount),rd))

# Checking crop dmg. exp
CDE <- data %>%
  count(CROPDMGEXP, name='CDEcount') %>%
  mutate (CDE_pct = round(100*CDEcount/sum(CDEcount),rd))

print(PDE, row.names= NULL)
print(CDE, row.names= NULL)
```

Although the erroneous EXP values for both EXP components were hardly significant it still had to be cleaned up to calculate the amount of damage.  

The erroneous EXP values were converted according to the following assumptions in order:  

1. All values were capitalized.  
2. All special characters were converted to *blanks*.  
3. All *blanks* were then converted to 0.  
4. Digits were considered as 1e+^(digit)^ multipliers.  

Property & crop damage values were then calculated and combined into variables "PropDmg" & "CropDmg" respectively in the  clean data set.

```{r Clean EXP}
# Replace erroneous values of Property & Crop Damages EXP.

## Cleaning prop & crop dmg. exp
data$PROPDMGEXP <- gsub("[^[:alnum:]]","",data$PROPDMGEXP)
data$CROPDMGEXP <- gsub("[^[:alnum:]]","",data$CROPDMGEXP)
data <-data %>%
  mutate(PROPDMGEXP = toupper(PROPDMGEXP),CROPDMGEXP = toupper(CROPDMGEXP)) %>%
  mutate(PROPDMGEXP = replace(PROPDMGEXP,PROPDMGEXP=="",0),
         CROPDMGEXP = replace(CROPDMGEXP,CROPDMGEXP=="",0)) %>%
  mutate(PROPDMGEXP = replace(PROPDMGEXP,PROPDMGEXP=='H', 100),
         CROPDMGEXP = replace(CROPDMGEXP,CROPDMGEXP=='H', 100)) %>%
  mutate(PROPDMGEXP = replace(PROPDMGEXP,PROPDMGEXP=='K', 1000),
         CROPDMGEXP = replace(CROPDMGEXP,CROPDMGEXP=='K', 1000)) %>%
  mutate(PROPDMGEXP = replace(PROPDMGEXP,PROPDMGEXP=='M', 1000000),
         CROPDMGEXP = replace(CROPDMGEXP,CROPDMGEXP=='M', 1000000)) %>%
  mutate(PROPDMGEXP = replace(PROPDMGEXP,PROPDMGEXP=='B', 1000000000),
         CROPDMGEXP = replace(CROPDMGEXP,CROPDMGEXP=='B', 1000000000)) %>%
  mutate(PROPDMGEXP = as.double(PROPDMGEXP),CROPDMGEXP = as.double(CROPDMGEXP))

## Calculating Damage Values and combining into new variables.
cleandata <-  data %>%
  mutate(PropDmg = PROPDMG*PROPDMGEXP, CropDmg = CROPDMG*CROPDMGEXP) %>%
  select("EVTYPE", "FATALITIES", "INJURIES", "PropDmg", "CropDmg" )
head(cleandata)
```
Showing top of the clean data set.  

## Results
### Analysis Logic
The clean data set was filtered and grouped to show the required categorical summaries.
Only the top 5 events in each category were plotted accordingly.  

Initial options for the plots:

```{r Plot Options}
# Options for Figures
mytheme <- ttheme_default(base_size=10,
                          core= list(fg_params= list(hjust=0,x=0.1)),
                          colhead= list(fg_params=list(hjust=0, x=0.1)))
top <-5
```

### Top 5 Weather Events Causing Fatalities & Injuries.

```{r Fatalities & Injuries}
## Fatalities
fatal <- cleandata %>%
  group_by(EVTYPE) %>%
  summarise(totfatal = sum(FATALITIES)) %>%
  arrange(desc(totfatal)) %>%
  filter(totfatal !=0) %>%
  mutate(pct=round(100*totfatal/sum(totfatal),1)) %>%
  slice_max(totfatal, n=top)

p1f <- ggplot(fatal,aes(x=totfatal, y=reorder(EVTYPE, totfatal)))
p1f <- p1f + geom_col(fill ='red')
p1f <- p1f + labs(title= paste('Fatal (Top',top, '=', sum(fatal$pct),'%)'),
                  y=NULL, x= 'Count')
fatal$totfatal <- format(fatal$totfatal, big.mark=",",core.just= "right")
tb1 <- tableGrob(fatal, theme= mytheme, rows = NULL, cols= c("Event Type","Fatalities","%"))

## Injured
injured <- cleandata %>%
  group_by(EVTYPE) %>%
  summarise(totinjured = sum(INJURIES)) %>%
  arrange(desc(totinjured)) %>%
  filter(totinjured !=0) %>%
  mutate(pct=round(100*totinjured/sum(totinjured),1)) %>%
  slice_max(totinjured,n=top)

p2i <- ggplot(injured, aes(y=reorder(EVTYPE, totinjured), x=totinjured))
p2i <- p2i + geom_col(fill ='orange')
p2i <- p2i + labs(title = paste('Injured (Top',top, '=', sum(fatal$pct),'%)'),
                  y=NULL, x= 'Count')
injured$totinjured <- format(injured$totinjured, big.mark=",",core.just= "right")
tb2 <- tableGrob(injured, theme= mytheme, rows = NULL, cols= c("Event Type","Injured","%"))

grid.arrange(p1f, p2i, tb1, tb2, heights= unit(c(120,80), c("mm", "mm")), widths= unit(c(120,120), c("mm", "mm")))
```

### Top 5 Weather Events Causing Damage.  

```{r Damages}
PDmg <- cleandata %>%
  group_by(EVTYPE) %>%
  summarise(totPDmg = round(sum(PropDmg)/1e9,3)) %>%
  arrange(desc(totPDmg)) %>%
  filter(totPDmg !=0) %>%
  mutate(pct=round(100*totPDmg/sum(totPDmg),1)) %>%
  slice_max(totPDmg,n=top)

p3p <- ggplot(PDmg, aes(y=reorder(EVTYPE, totPDmg), x= totPDmg))
p3p <- p3p + geom_col(fill ='Purple')
p3p <- p3p + labs(title= paste('Property Damage (Top', top, '=', sum(PDmg$pct),'%)'),
                  y= NULL, x= 'Value ($B)')
PDmg$totPDmg <- format(PDmg$totPDmg, big.mark=",",core.just= "right")
tb3 <- tableGrob(PDmg, theme= mytheme, rows = NULL, cols= c("Event Type","Value ($B)","%"))

## Crop Damage
CDmg <- cleandata %>%
  group_by(EVTYPE) %>%
  summarise(totCDmg = round(sum(CropDmg)/1e9,3)) %>%
  arrange(desc(totCDmg)) %>%
  filter(totCDmg !=0) %>%
  mutate(pct=round(100*totCDmg/sum(totCDmg),1)) %>%
  slice_max(totCDmg,n=top)

p4c <- ggplot(CDmg,aes(y=reorder(EVTYPE, totCDmg), x=totCDmg))
p4c <- p4c + geom_col(fill ='Steel Blue')
p4c <- p4c + labs(title= paste('Crop Damage (Top',top, '=', sum(CDmg$pct),'%)'),
                  y= NULL, x= 'Value ($B)')
CDmg$totCDmg <- format(CDmg$totCDmg, big.mark=",",core.just= "right")
tb4 <- tableGrob(CDmg, theme= mytheme, rows = NULL, cols= c("Event Type","Value ($B)","%"))

grid.arrange(p3p,p4c, tb3, tb4, heights= unit(c(120,80), c("mm", "mm")), widths= unit(c(120,120), c("mm", "mm")))
```
